---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# weathR R package

## Written by Roisin Murphy, Alanna Collier, Sean Carroll and Jack Castles. 
<!-- badges: start -->
<!-- badges: end -->

## Description

The goal of weathR is to provide tools for loading, cleaning, analysing, and visualising climate data taken from Dublin airport throughout 1981-2010. The package includes functions to preprocess climate data, compute seasonal statistics (e.g., mean, median), and visualise trends interactively.

## Installation

You can install the development version of weathR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("seancarrolll/weathR")
```

## Example

This function loads the Dublin Airport dataset and ensures that it is properly formatted and classified as a "dublinairport" object.
It then displays the first 6 rows:

```{r example}
library(weathR)
dublinairport <- load_dublinairport()
print(head(dublinairport))
```

```{r}
summarise.dublinairport(dublinairport)
```

Subsequently, an animated line plot visualising trends in temperature, daily sunshine, and rainfall over the months can be created using the following function:
 
```{r animation example, warning=FALSE, messages= FALSE}
 #anim <- plot_summarise(dublinairport)
 # print(anim)
```

A more through introduction can be found in the `weathR` vignette.
