---
title: "weathR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{weathR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>,"
)
```

```{r setup}
if (!requireNamespace("weathR", quietly = TRUE)) {
  message("The 'weathR' package is not installed. Using mock data for demonstration.")
} else {
  library(weathR)
}
```
  
##INTRODUCTION
The weathR package is designed for analysing and visualising seasonal climate
data. It provides tools for loading datasets, computing summary statistics,
and creating animated visualisations of climate trends.

This vignette walks through key functionalities of the package, including
loading the built-in Dublin Airport Climate Averages dataset, summarising data, 
and visualising trends


##INSTALLATION
Once in R you can type at the R command below to install weathR
# install.packages("pak")
pak::pak("seancarrolll/weathR")

##Loading the Dataset
The weathR package comes with a built-in dataset: the Dublin Airport Climate
Averages (1981–2010). To load this dataset, use the load_dublinairport() function:

```{r}
if (!requireNamespace("weathR", quietly = TRUE)) {
  message("Using a mock dataset for demonstration.")
  
  # Mock dataset
  climate_data <- data.frame(
    Month = factor(
      c("January", "February", "March", "April", "May", "June",
        "July", "August", "September", "October", "November", "December"),
      levels = month.name
    ),
    Temperature = c(5.0, 5.5, 7.0, 9.0, 12.0, 15.0, 17.0, 16.5, 14.0, 10.5, 7.5, 5.5)
  )
} else {
  library(weathR)
  
  # Load the dataset
  climate_data <- load_dublinairport()
}

# Display the first few rows of the dataset
head(climate_data)

```

#Summarising Data
The summarise.dublinairport() function allows you to compute summary statistics
for the dataset, including mean, median, and quantiles. You can optionally 
filter the dataset by season or month.
```{r}
if (exists("summarise.dublinairport")) {
  # Summarize the dataset using weathR
  summary_table <- summarise.dublinairport(climate_data)
} else {
  #Mock summary table for demonstration
  summary_table <- data.frame(
    Statistic = c("Mean Temperature"),
    Value = c(mean(climate_data$Temperature, na.rm = TRUE))
  )
}

#Display the summary table
print(summary_table)
```
#Data Peparation for Plotting
Next is to reshape the data so that you are able to fit it into a suitable plot. 
```{r}
# Reshape data for plotting
data_long <- tidyr::pivot_longer(
 climate_data,
 cols = c("Temperature"),
 names_to = "Variables",
 values_to = "Value"
)
```

#Creating Visualisations
Static Visualisation
You can create static plots of climate data trends using ggplot2. For example,
a line plot of monthly average temperatures can be created like this:
 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
#Create Static plot with temperature variable
p <- ggplot(data_long, aes(x = Month, y = Value, color = Variables, group = Variables)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Monthly Temperature Trends",
    x = "Month",
    y = "Temperature (°C)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p
```

#Animated Visualisation
The plot_summarise() function generates an animated visualisation of seasonal
climate trends using gganimate and ggplot2:
```{r ,warning=FALSE, message=FALSE}

library(gganimate)
# Add animation effects
p_animated <- p +
  transition_reveal(as.numeric(Month)) +
  ease_aes('linear')

# Render the animation
animation <- animate(
  p_animated,
  duration = 10,
  fps = 20,
  width = 800,
  height = 600,
  renderer = gifski_renderer(),
  nframes = 200
)

# Display the animation 
animation
```


